Notes on Validated Model Counting \\ Version of \today

### 2022-08-05 doc/mc-notes.tex
Use POG as name of graph representation

For ring elemement $a$, we use the notation $\oneminus a$ to denote $1 - a$.

@@

\section{Separable Schemas}

Computing the cost function for a Boolean formula becomes
straightforward when the formula contains only the operations
$\pand$, $\por$, and $\neg$, as is demonstrated by
Propositions~\ref{prop:negation}--\ref{prop:disjunction}.  We define {\em separable schemas}
as a direct-acyclic graph representation of such formulas.
Representing formulas as a graph enables sharing subformulas, yielding a more compact representation.

\subsection{Schema Definition}

\begin{table}
  \caption{Recursive Definition of Separable Schemas and their Costs}
  \label{tab:schema}
%%A) Basic Operations\\
\begin{center}
\begin{tabular}{ccccc}
\toprule
  \makebox[15mm]{$S$}  & \makebox[30mm]{Restrictions} & \makebox[30mm]{$D(S)$} & \makebox[30mm]{$\interpset{S}$} & \makebox[25mm]{$\cost(S)$}\\
\midrule
  $\nil$  & None  & $\emptyset$    & $\emptyset$  & $0$\\
  $\tautology$        & None  & $\emptyset$    & $\uinterp$ & $1$ \\
$x_i$           & None  & $\{ x_i \}$    & $\{ \interp | \interp(x_i) = \tautology \}$ & $\cost(x_i)$ \\
$\neg S_1$      & None  & $\depend(S_1)$ & $\uinterp - \interpset{S_1}$ & $\oneminus\cost(S)$ \\
$S_1 \pand S_2$ & $\depend(S_1) \cap \depend(S_2) = \emptyset$ & $\depend(S_1) \cup \depend(S_2)$ & $\interpset{S_1} \cap \interpset{S_2}$
& $\cost(S_1) \cdot \cost(S_2)$\\
$S_1 \por S_2$  & $\interpset{S_1} \cap \interpset{S_2} = \emptyset$ &  $\depend(S_1) \cup \depend(S_2)$ &  $\interpset{S_1} \cup \interpset{S_2}$
& $\cost(S_1) + \cost(S_2)$\\
\bottomrule
\end{tabular}
\end{center}
%% B) Composite Operations\\
%% \begin{center}
%% \begin{tabular}{cc}
%% \toprule
%%   $S$           & Definition \\
%% \midrule
%% \makebox[30mm]{$\pite(S_1, S_2, S_3)$} & \makebox[40mm]{$[S_1 \pand S_2] \por [\neg S_1 \pand S_2]$}\\
%% \bottomrule
%% \end{tabular}
%% \end{center}
\end{table}

The set of schemas over a set of variables
$\{x_1, x_2, \ldots, x_n\}$
can be defined recursively, as is shown in Table
\ref{tab:schema}.  Each schema $S$ has an associated dependency set
$\depend(S)$ and an associated set of models $\interpset{S}$.
%% We
%% divide the operations into {\em basic} ones that are fundamental to
%% the representation, and {\em composite} ones that can be constructed
%% from basic operations.  The latter category consists of just the $\pite$ operation.
A key property of a Boolean formula represented by separable schema
$S$ is that, for any separable cost function $\cost$, the cost of the
formula $\cost(S)$ can be computed with a linear number of ring
operations, as is shown in the final column of Table~\ref{tab:schema}.

\subsection{Schema Normalization}

\begin{table}
  \caption{Normalization Rules}
  \label{tab:normalize}
  \begin{center}
  \begin{tabular}{ccccccc}
    \makebox[20mm]{$\neg \nil$} & \makebox[5mm]{$\rightarrow$} & \makebox[20mm]{$\tautology$} & \makebox[15mm]{} &
    \makebox[20mm]{$\neg \tautology$} & \makebox[5mm]{$\rightarrow$} & \makebox[20mm]{$\nil$} \\
    $\neg \neg S$ & $\rightarrow$ & $S$ \\
    $S \pand \nil$ & $\rightarrow$ & $\nil$ && $\nil \pand S$ & $\rightarrow$ & $\nil$ \\
    $S \pand \tautology$ & $\rightarrow$ & $S$ && $\tautology \pand S$ & $\rightarrow$ & $S$ \\
    $S \por \nil$ & $\rightarrow$ & $S$ && $\nil \por S$ & $\rightarrow$ & $S$ \\
    $S \por \tautology$ & $\rightarrow$ & $\tautology$ && $\tautology \por S$ & $\rightarrow$ & $\tautology$ \\
%%     $\pite(\neg S_1, S_2, S_3)$ & $\rightarrow$ & $\pite(S_1, S_3, S_2)$ \\
%%     $\pite(\tautology, S_2, S_3)$ & $\rightarrow$ & $S_2$ & &
%%     $\pite(\nil, S_2, S_3)$ & $\rightarrow$ & $S_3$ \\
%%     $\pite(S_1, \tautology, \nil)$ & $\rightarrow$ & $S_1$ &&
%%     $\pite(S_1, \nil, \tautology)$ & $\rightarrow$ & $\neg S_1$ \\
%%     $\pite(S_1, S_2, \nil)$ & $\rightarrow$ & $S_1 \pand S_2$ & &
%%     $\pite(S_1, \nil, S_3)$ & $\rightarrow$ & $\neg S_1 \pand S_3$ \\
%%     $\pite(S_1, S_2, \tautology)$ & $\rightarrow$ & $\neg [S_1 \pand \neg S_2]$ & &
%%     $\pite(S_1, \tautology, S_3)$ & $\rightarrow$ & $\neg [\neg S_1 \pand \neg S_3]$ \\
  \end{tabular}
  \end{center}
\end{table}

Table~\ref{tab:normalize} shows a list of {\em normalizing}
transformations to simplify a separable schema.  These 
eliminate extra negations and remove constant
terms, such that constants only occur in schemas when representing constant functions $\zero$ and $\one$.
%% The most interesting are the two bottom rules, which
%% use DeMorgan's Laws to simplify an $\pite$ operation when one of
%% the arguments is $\tautology$ into a $\pand$ operation.  We will make use of this
%% transformation when using BDDs to convert a set of clauses into a
%% separable schema.

@@

The if-then-else (ITE) operation arises when converting the CNF
representation of a formula into a separable schema, both for
bottom-up approaches based on decision diagrams, and for top-down
approaches based on CDCL\@.  For functions $f_1$, $f_2$, and $f_3$, we
define
$\ite(f_1, f_2, f_3) = (f_1 \land f_2) \lor (\neg f_1 \land f_3)$.
Observe that the $\lor$ operation in this expression satisfies
the requirements for $\por$, since the first argument can only yield
$\tautology$ for assignments that yield $\tautology$ for $f_1$, while
the second can only yield $\tautology$ for assignments that yield
$\nil$ for $f_1$.  Therefore, the only condition imposed on an
expansion of $\ite$ into the allowed schema operations is that the
dependency set for $f_1$ must be disjoint from those of $f_2$ and
$f_3$.

\begin{table}
  \caption{Encodings of the ITE Operation}
  \label{tab:ite}
  \begin{center}
  \begin{tabular}{ccc}
    ITE Form & & Encoding \\
    \midrule
     $\ite(S_1, S_2, S_3)$ & $$ & $(S_1 \pand S_2) \por (\neg S_1 \land S_3)$ \\
     $\ite(\tautology, S_2, S_3)$ & $$ & $S_2$ \\
     $\ite(\nil, S_2, S_3)$ & $$ & $S_3$ \\
     $\ite(S_1, \tautology, \nil)$ & $$ & $S_1$ \\
     $\ite(S_1, \nil, \tautology)$ & $$ & $\neg S_1$ \\
     $\ite(S_1, S_2, \nil)$ & $$ & $S_1 \pand S_2$ \\
     $\ite(S_1, \nil, S_3)$ & $$ & $\neg S_1 \pand S_3$ \\
     $\ite(S_1, \tautology, S_3)$ & $$ & $\neg (\neg S_1 \pand \neg S_3)$ \\
     $\ite(S_1, S_2, \tautology)$ & $$ & $\neg (S_1 \pand \neg S_2)$ \\
  \end{tabular}
  \end{center}
\end{table}

Table~\ref{tab:ite} shows different ways to encode an ITE operation
into schema operations.  All of these require the dependency set of
the argument $S_1$ to be disjoint from those of arguments $S_2$ and
$S_3$.  The first row shows the most general case, requiring one
$\por$ and two $\pand$ operations.  The other rows show special cases,
where one or more argument is a constant.  Of these, the final two
rows are particularly noteworthy.  They make use of DeMorgan's Laws to
convert disjunctions into conjunctions.  In particular, for Boolean
functions $f_1$ and $f_2$, we can write $\ite(f_1, \tautology, f_2)$
as $f_1 \lor f_2$, and by DeMorgan's Laws as $\neg(\neg f_1 \land
\neg f_2)$.  Similarly $\ite(f_1, f_2, \tautology) = \neg f_1 \lor f_2 = \neg(f_1 \land \neg f_2)$.
These conjunctions can then be
encoded with $\pand$ operations, since their arguments will have
disjoint dependency sets.

@@

The CRAT clausal proof framework provides a means to express a
checkable proof that a Boolean formula, given in conjunctive normal
form, is logically equivalent to a separable schema.  Once this
equivalence has been established, the schema can form the basis
for computations enabled by the
representation, including trusted model counting

@@

Table~\ref{tab:crat:syntax} shows the set of proof rules for the CRAT
format.
As with other clausal proof formats, a variable is
represented by a positive integer $v$, with the first ones being input
variables and successive ones being extension variables.  Literal $l$
is represented by a signed integer, with $-v$ being the complement of
variable $v$.  Each clause is indicated by a positive integer
identifier $C$, with the first ones being the IDs of the input clauses and successive
ones being the IDs of added clauses.  Clause identifiers must be totally ordered,
such that clause $C$ can only reference clauses $C'$ such that $C' < C$.
Clause identifiers need not be consecutive.

@@

A CRAT proof follows the same general form as a QRAT dual
proof~\cite{bryant:cade:2021}---one that ensures that each clause
addition and each clause deletion preserves equivalence.  With CRAT,
however, clauses are defined both explicitly and implicitly.  Starting
with the set of input clauses, the proof consists of a sequence of steps that
both add and delete clauses.  Each addition must be truth preserving
and each deletion must be falsehood preserving.  At the end, all input
clauses must have been deleted, and among the remaining clauses there
must be only a single unit clause consisting of some variable or its
complement.  Except for trivial cases, the final literal will be an
extension variable or its complement.  That literal will indicate the
root of the schema as the (possibly negated) output of a schema
operation.  Working from that root backward, the schema can be
extracted from the CRAT file.

@@

\begin{figure}
  \centering{
    \begin{minipage}{0.4\textwidth}
      A) Schema   \\[2ex]
    \centering{\input{dd/c3}}
    \end{minipage}
    \begin{minipage}{0.4\textwidth}
    B) Cost Computation\\[2ex]
    \centering{\input{dd/c3-eval}}
    \end{minipage}
  }
\caption{Schema \#1 for Formula $\phi_1 = x_1 \lor x_2 \lor x_3$ and its Cost Computation}
\label{fig:c3:schema}
\end{figure}

@@

As an illustration, consider the Boolean formula
$\phi_1 = x_1 \lor x_2 \lor x_3$,
represented by a single clause.  We cannot directly use the
$\por$ operation to form these disjunctions, since the sets of assignments
satisfying the individual literals are not disjoint.  Instead, we must
decompose this formula into a sequence of operations.
Figure~\ref{fig:c3:schema}A shows one such decomposition.
The subscripts of the variables and the operator labels correspond to
the numbers of the input and extension variables in the CRAT proof.
Edges marked with dots indicate Boolean negation.

@@

The logical representation can readily be converted into a formula for
computing $\cost(\phi_1)$, the cost of formula $\phi_1$, given a weight
$w_i$ for each variable $x_i$ for $1 \leq i \leq 3$.  This is
illustrated in Figure~\ref{fig:c3:schema}B\@.  Note how the Boolean
negations become $\oneminus$ operations.  This formula is valid for
any cost function.

@@

\begin{figure}
  \centering{
    \begin{minipage}{0.4\textwidth}
      A) Schema   \\[2ex]
    \centering{\input{dd/p2-bdd}}
    \end{minipage}
    \begin{minipage}{0.4\textwidth}
    B) Cost Computation\\[2ex]
    \centering{\input{dd/p2-bdd-eval}}
    \end{minipage}
  }
  \caption{Schema \#2A for Formula $\phi_2 = (x_1 \lor x_2 \lor x_3) \land (\obar{x}_1 \lor x_2)$ and its Cost Computation}
\label{fig:p2-bdd:schema}
\end{figure}

@@

Figure~\ref{fig:p2-bdd:schema} shows a schema representing the formula,
and Figure~\ref{fig:p2-bdd:crat} shows the associated CRAT file and proof clauses.  This
proof was generated via a bottom-up strategy, such as would be created
using BDDs.  It creates schematic representations of the input
clauses and then forms their conjunction.  In the proof, unit
clauses are generated for the representations of the input clauses,
and then the input clauses are deleted.  These intermediate unit
clauses are used to justify a unit clause for the final root, and then
they are deleted.

Using DeMorgan's Laws, $C_1$ can be written as 
$t_1 = \neg [\obar{x}_1\land \obar{x}_2\land \obar{x}_3]$, and this can be
expressed using the $\pand$ operation $p_4$, shown in
Figure~\ref{fig:p2-bdd:schema}A\@.  (Note that $t_1$ in this case is
logically equivalent to root $t_1$ in the schema of
Figure~\ref{fig:c3:schema}A, but the use of negation enables it to be
represented with one operation rather than four.)  Similarly, $C_2$
can be written as $t_2 = \neg [x_1\land \obar{x}_2]$, and this can be
represented by $\pand$ operation $p_5$.
Terms $t_1$ and $t_2$ are asserted as unit clauses on proof lines 7 and 11,
allowing the input clauses to be deleted.
The conjunction $t_1 \land t_2$
can be written as $t_3 = \neg[\obar{t_1} \lor \obar{t_2}]$, represented by $\por$ operation $s_6$.
Term $t_3$ is asserted as a unit clause on proof line 15.
Based on this, the unit clauses for terms $t_1$ and
$t_2$ can be deleted.  Term $t_3$ then becomes the unique root of the cost
computation shown in Figure~\ref{fig:p2-bdd:schema}B\@.

@@

\begin{figure}
  \centering{
    \begin{minipage}{0.4\textwidth}
      A) Schema   \\[2ex]
    \centering{\input{dd/p2-cdcl}}
    \end{minipage}
    \begin{minipage}{0.4\textwidth}
    B) Cost Computation\\[2ex]
    \centering{\input{dd/p2-cdcl-eval}}
    \end{minipage}
  }
\caption{Schema \#2B for Formula $\phi_2 = (x_1 \lor x_2 \lor x_3) \land (\obar{x}_1 \lor x_2)$ and its Cost Computation}
\label{fig:p2-cdcl:schema}
\end{figure}

@@

Figure~\ref{fig:p2-cdcl:schema} shows an alternate schema representing
the same formula $\phi_2$ as in Example~2A, and
Figure~\ref{fig:p2-cdcl:crat} shows the associated CRAT proof.  This
proof was generated via a top-down strategy, such as would be created
using a model counter based on CDCL\@.  It starts by splitting on
variable $x_1$.  Clause $C_1$ is trivially satisfied when $x_1$ is
assigned $\tautology$, and clause $C_2$ becomes the clause $C'_2 =
x_2$.  On the other hand, clause $C_2$ is trivially satisfied when
$x_1$ is assigned $\nil$, and clause $C_1$ becomes $C'_1 = x_2 \lor
x_3$.  Clause $C'_1$ can be represented schematically as
$\neg[\obar{x}_2 \pand \obar{x}_3]$, with this operation labeled $p_4$
in Figure~\ref{fig:p2-cdcl:schema}A\@.  The splitting on variable
$x_1$ can be rejoined as $\ite(x_1, x_2, \obar{p}_4)$, and this ITE
operation can be expressed using the product operations $p_5$ and
$p_6$, joined by the sum operation $t = s_7$.

The two schemas shown in Figures~\ref{fig:p2-bdd:schema}A and
\ref{fig:p2-cdcl:schema}A have similar structure.  They have very
different negation patterns, but they are logically equivalent.  Their
associated cost computations (Figures~\ref{fig:p2-bdd:schema}B and
\ref{fig:p2-cdcl:schema}B) also yield the same results for arbitrary
weights $w_1$, $w_2$, and $w_3$.

@@

Given an arbitrary CNF formula, we can use BDD operations to generate
a schematic representation.  The proof generation can follow the
methods we have used for generating unsatisfiability proofs of Boolean
formulas~\cite{bryant:tacas:2021} and dual proofs of quantified
Boolean formulas~\cite{bryant:cade:2021}.
Each BDD node can be expressed as an ITE operation and
make use of the encodings shown in Table~\ref{tab:ite}.

A second class of model counters proceeds top-down, based on the CDCL
framework.  These choose a splitting variable $x$ and recursively
construct schemas $S_1$ and $S_0$ for the two assignments to the
variable.  These are combined as $\ite(x, S_1, S_0)$, using one of the
encodings of $\ite$ shown in Table~\ref{tab:ite}.  Like CDCL, the
top-down algorithm can make use of unit propagation, conflict
detection, and clause learning.  It can also make use of {\em
  variable} partitioning.  That is, suppose for some partial
assignment to the variables, the input clauses decompose into two or
more sets over disjoint variables.  Then the schema for each of these
partitions can be generated separately, and these are joined via the $\pand$
operation.

>>>>

For ring element $a$, we use the notation $\oneminus a$ to denote $1 - a$.

@@

\section{Partitioned-Operation Graphs (POGs)}

Computing the cost function for a Boolean formula becomes
straightforward when the formula contains only the operations
$\pand$, $\por$, and $\neg$, as is demonstrated by
Propositions~\ref{prop:negation}--\ref{prop:disjunction}.  We define a {\em partitioned-operation graph} (POG)
as a direct-acyclic graph representation of such formulas.
Representing formulas as a graph enables sharing subformulas, yielding a more compact representation.

\subsection{POG Definition}

\begin{table}
  \caption{Recursive Definition of POGs and their Costs}
  \label{tab:pog}
%%A) Basic Operations\\
\begin{center}
\begin{tabular}{ccccc}
\toprule
  \makebox[15mm]{$P$}  & \makebox[30mm]{Restrictions} & \makebox[30mm]{$D(P)$} & \makebox[30mm]{$\interpset{P}$} & \makebox[25mm]{$\cost(P)$}\\
\midrule
  $\nil$  & None  & $\emptyset$    & $\emptyset$  & $0$\\
  $\tautology$        & None  & $\emptyset$    & $\uinterp$ & $1$ \\
$x_i$           & None  & $\{ x_i \}$    & $\{ \interp | \interp(x_i) = \tautology \}$ & $\cost(x_i)$ \\
$\neg P_1$      & None  & $\depend(P_1)$ & $\uinterp - \interpset{P_1}$ & $\oneminus\cost(P)$ \\
$P_1 \pand P_2$ & $\depend(P_1) \cap \depend(P_2) = \emptyset$ & $\depend(P_1) \cup \depend(P_2)$ & $\interpset{P_1} \cap \interpset{P_2}$
& $\cost(P_1) \cdot \cost(P_2)$\\
$P_1 \por P_2$  & $\interpset{P_1} \cap \interpset{P_2} = \emptyset$ &  $\depend(P_1) \cup \depend(P_2)$ &  $\interpset{P_1} \cup \interpset{P_2}$
& $\cost(P_1) + \cost(P_2)$\\
\bottomrule
\end{tabular}
\end{center}
%% B) Composite Operations\\
%% \begin{center}
%% \begin{tabular}{cc}
%% \toprule
%%   $P$           & Definition \\
%% \midrule
%% \makebox[30mm]{$\pite(P_1, P_2, P_3)$} & \makebox[40mm]{$[P_1 \pand P_2] \por [\neg P_1 \pand P_2]$}\\
%% \bottomrule
%% \end{tabular}
%% \end{center}
\end{table}

The set of POGs over a set of variables
$\{x_1, x_2, \ldots, x_n\}$
can be defined recursively, as is shown in Table
\ref{tab:pog}.  Each POG $P$ has an associated dependency set
$\depend(P)$ and an associated set of models $\interpset{P}$.
%% We
%% divide the operations into {\em basic} ones that are fundamental to
%% the representation, and {\em composite} ones that can be constructed
%% from basic operations.  The latter category consists of just the $\pite$ operation.
A key property of a Boolean formula represented by POG
$P$ is that, for any separable cost function $\cost$, the cost of the
formula $\cost(P)$ can be computed with a linear number of ring
operations, as is shown in the final column of Table~\ref{tab:pog}.

\subsection{POG Normalization}

\begin{table}
  \caption{Normalization Rules}
  \label{tab:normalize}
  \begin{center}
  \begin{tabular}{ccccccc}
    \makebox[20mm]{$\neg \nil$} & \makebox[5mm]{$\rightarrow$} & \makebox[20mm]{$\tautology$} & \makebox[15mm]{} &
    \makebox[20mm]{$\neg \tautology$} & \makebox[5mm]{$\rightarrow$} & \makebox[20mm]{$\nil$} \\
    $\neg \neg P$ & $\rightarrow$ & $P$ \\
    $P \pand \nil$ & $\rightarrow$ & $\nil$ && $\nil \pand P$ & $\rightarrow$ & $\nil$ \\
    $P \pand \tautology$ & $\rightarrow$ & $P$ && $\tautology \pand P$ & $\rightarrow$ & $P$ \\
    $P \por \nil$ & $\rightarrow$ & $P$ && $\nil \por P$ & $\rightarrow$ & $P$ \\
    $P \por \tautology$ & $\rightarrow$ & $\tautology$ && $\tautology \por P$ & $\rightarrow$ & $\tautology$ \\
%%     $\pite(\neg P_1, P_2, P_3)$ & $\rightarrow$ & $\pite(P_1, P_3, P_2)$ \\
%%     $\pite(\tautology, P_2, P_3)$ & $\rightarrow$ & $P_2$ & &
%%     $\pite(\nil, P_2, P_3)$ & $\rightarrow$ & $P_3$ \\
%%     $\pite(P_1, \tautology, \nil)$ & $\rightarrow$ & $P_1$ &&
%%     $\pite(P_1, \nil, \tautology)$ & $\rightarrow$ & $\neg P_1$ \\
%%     $\pite(P_1, P_2, \nil)$ & $\rightarrow$ & $P_1 \pand P_2$ & &
%%     $\pite(P_1, \nil, P_3)$ & $\rightarrow$ & $\neg P_1 \pand P_3$ \\
%%     $\pite(P_1, P_2, \tautology)$ & $\rightarrow$ & $\neg [P_1 \pand \neg P_2]$ & &
%%     $\pite(P_1, \tautology, P_3)$ & $\rightarrow$ & $\neg [\neg P_1 \pand \neg P_3]$ \\
  \end{tabular}
  \end{center}
\end{table}

Table~\ref{tab:normalize} shows a list of {\em normalizing}
transformations to simplify a POG\@.  These 
eliminate extra negations and remove constant
terms, such that constants only occur in POGs when representing constant functions $\zero$ and $\one$.
%% The most interesting are the two bottom rules, which
%% use DeMorgan's Laws to simplify an $\pite$ operation when one of
%% the arguments is $\tautology$ into a $\pand$ operation.  We will make use of this
%% transformation when using BDDs to convert a set of clauses into a
%%  POG\@.

@@

The if-then-else (ITE) operation arises when converting the CNF representation
of a formula into a  POG, both for
bottom-up approaches based on decision diagrams, and for top-down
approaches based on CDCL\@.  For functions $f_1$, $f_2$, and $f_3$, we
define
$\ite(f_1, f_2, f_3) = (f_1 \land f_2) \lor (\neg f_1 \land f_3)$.
Observe that the $\lor$ operation in this expression satisfies
the requirements for $\por$, since the first argument can only yield
$\tautology$ for assignments that yield $\tautology$ for $f_1$, while
the second can only yield $\tautology$ for assignments that yield
$\nil$ for $f_1$.  Therefore, the only condition imposed on an
expansion of $\ite$ into the allowed POG operations is that the
dependency set for $f_1$ must be disjoint from those of $f_2$ and
$f_3$.

\begin{table}
  \caption{Encodings of the ITE Operation}
  \label{tab:ite}
  \begin{center}
  \begin{tabular}{ccc}
    ITE Form & & Encoding \\
    \midrule
     $\ite(P_1, P_2, P_3)$ & $$ & $(P_1 \pand P_2) \por (\neg P_1 \land P_3)$ \\
     $\ite(\tautology, P_2, P_3)$ & $$ & $P_2$ \\
     $\ite(\nil, P_2, P_3)$ & $$ & $P_3$ \\
     $\ite(P_1, \tautology, \nil)$ & $$ & $P_1$ \\
     $\ite(P_1, \nil, \tautology)$ & $$ & $\neg P_1$ \\
     $\ite(P_1, P_2, \nil)$ & $$ & $P_1 \pand P_2$ \\
     $\ite(P_1, \nil, P_3)$ & $$ & $\neg P_1 \pand P_3$ \\
     $\ite(P_1, \tautology, P_3)$ & $$ & $\neg (\neg P_1 \pand \neg P_3)$ \\
     $\ite(P_1, P_2, \tautology)$ & $$ & $\neg (P_1 \pand \neg P_2)$ \\
  \end{tabular}
  \end{center}
\end{table}

Table~\ref{tab:ite} shows different ways to encode an ITE operation
into POG operations.  All of these require the dependency set of
the argument $P_1$ to be disjoint from those of arguments $P_2$ and
$P_3$.  The first row shows the most general case, requiring one
$\por$ and two $\pand$ operations.  The other rows show special cases,
where one or more argument is a constant.  Of these, the final two
rows are particularly noteworthy.  They make use of DeMorgan's Laws to
convert disjunctions into conjunctions.  In particular, for Boolean
functions $f_1$ and $f_2$, we can write $\ite(f_1, \tautology, f_2)$
as $f_1 \lor f_2$, and by DeMorgan's Laws as $\neg(\neg f_1 \land
\neg f_2)$.  Similarly $\ite(f_1, f_2, \tautology) = \neg f_1 \lor f_2 = \neg(f_1 \land \neg f_2)$.
These conjunctions can then be
encoded with $\pand$ operations, since their arguments will have
disjoint dependency sets.

@@

The CRAT clausal proof framework provides a means to express a
checkable proof that a Boolean formula, given in conjunctive normal
form, is logically equivalent to a POG\@.  Once this
equivalence has been established, the POG can form the basis
for computations enabled by the
representation, including trusted model counting

@@

Table~\ref{tab:crat:syntax} shows the declarations that can occur in a CRAT file.
As with other clausal proof formats, a variable is
represented by a positive integer $v$, with the first ones being input
variables and successive ones being extension variables.  Literal $l$
is represented by a signed integer, with $-v$ being the complement of
variable $v$.  Each clause is indicated by a positive integer
identifier $C$, with the first ones being the IDs of the input clauses and successive
ones being the IDs of added clauses.  Clause identifiers must be totally ordered,
such that clause $C$ can only reference clauses $C'$ such that $C' < C$.
Clause identifiers need not be consecutive.

@@

A CRAT proof follows the same general form as a QRAT dual
proof~\cite{bryant:cade:2021}---one that ensures that each clause
addition and each clause deletion preserves equivalence.  With CRAT,
however, clauses are defined both explicitly and implicitly.  Starting
with the set of input clauses, the proof consists of a sequence of steps that
both add and delete clauses.  Each addition must be truth preserving
and each deletion must be falsehood preserving.  At the end, all input
clauses must have been deleted, and among the remaining clauses there
must be only a single unit clause consisting of some variable or its
complement.  Except for trivial cases, the final literal will be an
extension variable or its complement.  That literal will indicate the
root of the POG as the (possibly negated) output of a POG
operation.  Working from that root backward, the POG can be
extracted from the CRAT file.

@@

\begin{figure}
  \centering{
    \begin{minipage}{0.4\textwidth}
      A) POG   \\[2ex]
    \centering{\input{dd/c3}}
    \end{minipage}
    \begin{minipage}{0.4\textwidth}
    B) Cost Computation\\[2ex]
    \centering{\input{dd/c3-eval}}
    \end{minipage}
  }
\caption{POG \#1 for Formula $\phi_1 = x_1 \lor x_2 \lor x_3$ and its Cost Computation}
\label{fig:c3:pog}
\end{figure}

@@

As an illustration, consider the Boolean formula
$\phi_1 = x_1 \lor x_2 \lor x_3$,
represented by a single clause.  We cannot directly use the
$\por$ operation to form these disjunctions, since the sets of assignments
satisfying the individual literals are not disjoint.  Instead, we must
decompose this formula into a sequence of operations.
Figure~\ref{fig:c3:pog}A shows one such decomposition.
The subscripts of the variables and the operator labels correspond to
the numbers of the input and extension variables in the CRAT proof.
Edges marked with dots indicate Boolean negation.

@@

The logical representation can readily be converted into a formula for
computing $\cost(\phi_1)$, the cost of formula $\phi_1$, given a weight
$w_i$ for each variable $x_i$ for $1 \leq i \leq 3$.  This is
illustrated in Figure~\ref{fig:c3:pog}B\@.  Note how the Boolean
negations become $\oneminus$ operations.  This formula is valid for
any cost function.

@@

\begin{figure}
  \centering{
    \begin{minipage}{0.4\textwidth}
      A) POG   \\[2ex]
    \centering{\input{dd/p2-bdd}}
    \end{minipage}
    \begin{minipage}{0.4\textwidth}
    B) Cost Computation\\[2ex]
    \centering{\input{dd/p2-bdd-eval}}
    \end{minipage}
  }
  \caption{POG \#2A for Formula $\phi_2 = (x_1 \lor x_2 \lor x_3) \land (\obar{x}_1 \lor x_2)$ and its Cost Computation}
\label{fig:p2-bdd:pog}
\end{figure}

@@

Figure~\ref{fig:p2-bdd:pog} shows a POG representing the formula,
and Figure~\ref{fig:p2-bdd:crat} shows the associated CRAT file and proof clauses.  This
proof was generated via a bottom-up strategy, such as would be created
using BDDs.  It creates POG representations of the input
clauses and then forms their conjunction.  In the proof, unit
clauses are generated for the representations of the input clauses,
and then the input clauses are deleted.  These intermediate unit
clauses are used to justify a unit clause for the final root, and then
they are deleted.

Using DeMorgan's Laws, $C_1$ can be written as 
$t_1 = \neg [\obar{x}_1\land \obar{x}_2\land \obar{x}_3]$, and this can be
expressed using the $\pand$ operation $p_4$, shown in
Figure~\ref{fig:p2-bdd:pog}A\@.  (Note that $t_1$ in this case is
logically equivalent to root $t_1$ in the POG of
Figure~\ref{fig:c3:pog}A, but the use of negation enables it to be
represented with one operation rather than four.)  Similarly, $C_2$
can be written as $t_2 = \neg [x_1\land \obar{x}_2]$, and this can be
represented by $\pand$ operation $p_5$.
Terms $t_1$ and $t_2$ are asserted as unit clauses on proof lines 7 and 11,
allowing the input clauses to be deleted.
The conjunction $t_1 \land t_2$
can be written as $t_3 = \neg[\obar{t_1} \lor \obar{t_2}]$, represented by $\por$ operation $s_6$.
Term $t_3$ is asserted as a unit clause on proof line 15.
Based on this, the unit clauses for terms $t_1$ and
$t_2$ can be deleted.  Term $t_3$ then becomes the unique root of the cost
computation shown in Figure~\ref{fig:p2-bdd:pog}B\@.

@@

\begin{figure}
  \centering{
    \begin{minipage}{0.4\textwidth}
      A) POG   \\[2ex]
    \centering{\input{dd/p2-cdcl}}
    \end{minipage}
    \begin{minipage}{0.4\textwidth}
    B) Cost Computation\\[2ex]
    \centering{\input{dd/p2-cdcl-eval}}
    \end{minipage}
  }
\caption{POG \#2B for Formula $\phi_2 = (x_1 \lor x_2 \lor x_3) \land (\obar{x}_1 \lor x_2)$ and its Cost Computation}
\label{fig:p2-cdcl:pog}
\end{figure}

@@

Figure~\ref{fig:p2-cdcl:pog} shows an alternate POG representing
the same formula $\phi_2$ as in Example~2A, and
Figure~\ref{fig:p2-cdcl:crat} shows the associated CRAT proof.  This
proof was generated via a top-down strategy, such as would be created
using a model counter based on CDCL\@.  It starts by splitting on
variable $x_1$.  Clause $C_1$ is trivially satisfied when $x_1$ is
assigned $\tautology$, and clause $C_2$ becomes the clause $C'_2 =
x_2$.  On the other hand, clause $C_2$ is trivially satisfied when
$x_1$ is assigned $\nil$, and clause $C_1$ becomes $C'_1 = x_2 \lor
x_3$.  Clause $C'_1$ can be represented by POG operation
$\neg[\obar{x}_2 \pand \obar{x}_3]$, with this operation labeled $p_4$
in Figure~\ref{fig:p2-cdcl:pog}A\@.  The splitting on variable
$x_1$ can be rejoined as $\ite(x_1, x_2, \obar{p}_4)$, and this ITE
operation can be expressed using the product operations $p_5$ and
$p_6$, joined by the sum operation $t = s_7$.

The two POGs shown in Figures~\ref{fig:p2-bdd:pog}A and
\ref{fig:p2-cdcl:pog}A have similar structure.  They have very
different negation patterns, but they are logically equivalent.  Their
associated cost computations (Figures~\ref{fig:p2-bdd:pog}B and
\ref{fig:p2-cdcl:pog}B) also yield the same results for arbitrary
weights $w_1$, $w_2$, and $w_3$.

@@

Given an arbitrary CNF formula, we can use BDD operations to generate
a POG representation.  The proof generation can follow the
methods we have used for generating unsatisfiability proofs of Boolean
formulas~\cite{bryant:tacas:2021} and dual proofs of quantified
Boolean formulas~\cite{bryant:cade:2021}.
Each BDD node can be expressed as an ITE operation and
make use of the encodings shown in Table~\ref{tab:ite}.

A second class of model counters proceeds top-down, based on the CDCL
framework.  These choose a splitting variable $x$ and recursively
construct POGs $P_1$ and $P_0$ for the two assignments to the
variable.  These are combined as $\ite(x, P_1, P_0)$, using one of the
encodings of $\ite$ shown in Table~\ref{tab:ite}.  Like CDCL, the
top-down algorithm can make use of unit propagation, conflict
detection, and clause learning.  It can also make use of {\em
  variable} partitioning.  That is, suppose for some partial
assignment to the variables, the input clauses decompose into two or
more sets over disjoint variables.  Then the POG for each of these
partitions can be generated separately, and these are joined via the $\pand$
operation.
